Visual narratives of the Covid-19 pandemic

### 2022-09-02 Review/Response.qmd

---
title: "Reviewer Response"
output: html_document
---

# AE

::: callout
## Overall Evaluation
Overall largely
This paper provides an interesting overview of graphics used to portray data on the COVID 19 pandemic, largely from popular media. It attempts to use ideas and principles of graphics design and human factors
studies of graphical perception to evaluate the effectiveness of these displays for the public and policy
makers.

All in all, I think it largely succeeds in these aims and would be a suitable and interesting article for JDSSV readers. However, the two reviewers make substantial suggestions and comments which need to be considered before it is ready for publication. I recommend acceptance, following suitable revision (together with a reply from the authors to these reviews).
:::

Specific reviewer comments are addressed below, here, we address additional discussions brought up by the AE in response to the reviewers' comments.

::: callout
## Reviewer 2 - Principles

To the reviewer’s comment, I add that there needs to some clear statement regarding the match
between the design of a graphic and the capabilities of the intended audience in determining whether a
graphic display is successful for its intended purpose. This is more important than simple design rules.
:::

We have added a short paragraph addressing this in the data communication subsection of the introduction, and used the opportunity to state that we have primarily focused on charts intended for mass consumption rather than experts in public health or epidemiology.

::: callout
## Other comments: Proportional symbol maps

I was very surprised to see no mention or illustration of anamorphic maps or cartograms, where the area of each geographical unit is transformed to represent the count (cases/deaths). I strongly suggest including something on this.
:::

We have added a cartogram to the discussion of proportional symbol maps and choropleths. Thank you for catching an obvious omission on our part.

::: callout
### Other comments: Color scheme obstructions

Color scheme is mentioned only in regard to Fig 6, but there are other examples where the color scheme used could be noted, either positively or negatively.

We have added additional analysis of color where we feel it is useful; Figure 12 is one case where we chose not to mention color in part because I (Susan) can't tell if it's effective or not - I'm colorblind, and while I have trouble with streamgraphs in general, that is particularly true if color is actually meaningful.

::: callout
### Other comments: Uncertainty

Uncertainty: In the list of aspects mentioned in the middle of P 2, I am struck by the absence of mention
of uncertainty in visual representations of COVID. Clearly, the data are complex as they are, but
understanding their uncertainty is also a topic that could be mentioned or discussed.

We wholeheartedly agree - uncertainty should be mentioned in the introduction, though I can't think of any specific examples of addressing the uncertainty through visualization (as opposed to footnotes about data quality) in mainstream graphics which were circulated at the time. Still, that is a rather glaring omission from both our discussion and the collection of graphs we've accumulated as interesting.

::: callout
### Other comments: Temporal narrative

In Section 3, The temporal narrative, there could usefully be some discussion, and perhaps an illustration
of the use of dynamic charts like the moving bubble chart, to liberate the X axis from time, using
animation or an interactive slider.

We've added a moving bubble chart - again, this is an obvious omission. Thank you for bringing it to our attention. Interestingly, it was somewhat difficult to find a chart like this in a mainstream media outlet when searching through retrospectively. Part of that may be the fact that the web isn't great at keeping archival content, and many of these dashboards have been overwritten through successive modification. I suspect another part, however, is that moving bubble charts require a lot of effort to read and understand, and possibly were more common on social media and in the data vis community writ large than in media intended for the general public.

::: callout
### Other comments: standardization

More prominence could be given to the thoughtful standardization that went into some of these
displays to provide comparability of representation across global and local geographies and over
multiple waves of the pandemic. For example, in the early days cases and deaths were tied to calendar
dates giving time series charts with different origins. The simple device (I believe from Burn-Murdoch of
the FT) of making these all start when N cases had been observed was a game changer. There are
probably other examples.

There may be other examples, but this is a good one. We've added a comment to that effect in the discussion of time series data, but we also mention standardization w.r.t. area and population in the choropleth/cartogram section. While the topic probably deserves a more thorough treatment, it's not easy to slide it in and maintain the current structure of the paper, so we hope that this suffices.

@@

## Reviewer 1

::: callout
### How was the sample of publications selected? Under what criteria?

As it stands, the selection seems arbitrary and very limited. Why the Financial Times and The New York Times and not others? Why not analyzing visualizations from non-Western countries? If the authors want to talk about a “global narrative”, the selection of graphics should also be global. A comparative cross-cultural study would be pretty productive, as it might reveal similarities and differences of graphics from different regions.
:::

The authors are from the United States and Germany; while we participated in many discussions on global social media platforms such as Twitter, we do not have the global perspective (or even facility with sufficient languages) to undertake a review of a global sample of media relating to COVID. We have added a section which explicitly states how charts were selected and what our individual biases are with respect to geography and culture.

@@

I believe that the commentary on the graphics the authors chose would need to be more extensive and detailed. Just to give an example, they compare choropleth maps to proportional symbol maps, but they don’t reflect on the many shortcomings of the latter, such as the usual overlap between the scaled symbols. This might be related to space limits in the journal, but I think that this is a paper that deserves to be at least double its length to be truly comprehensive and detailed enough.

The authors also make too many assumptions about how a graphic will be interpreted. This is a direct quote: “The advantage of these displays is that they are very simple and allow for viewers to gain an intuitive understanding of the data”. Where is the evidence for such an assertion? The fact that the authors —likely professionals who are very familiar with statistics and visualization— find a graphic intuitive doesn’t make that graphic intuitive to other audiences. (The authors themselves actually address this in the following section, about log scales, arguably the best one in the paper.)

:::

While the editor has cautioned us not to double the length of the paper, where possible we have attempted to address some obvious shortcomings in our initial discussion without unduly increasing the paper's length. This paper is in part a result of discussing COVID graphics in several groups and seminars with data visualization professionals and also with laypeople; as a result, while we cannot directly cite a source saying that a specific graphic is simple and intuitive, we can directly assess the visual complexity of the chart and we can distill down the series of discussions we've had about these topics over the course of the pandemic to arrive at what we hope are fairly useful conclusions. In an attempt to be a bit more precise, however, where we could find relevant discussions on social media, we have included citations of those discussions to provide some support for claims that we have made summarizing the views of others.

@@

## Reviewer 2

::: callout
### Phrasing

This is clearly an article by researchers for whom English is likely a second language. While fluent, the wording is often obtuse and clunky. I have made a number of changes in the text to show where it might be improved.
:::

We appreciate the edits. We have addressed many of the stylistic suggestions where appropriate. In other cases, we have modified the paper in ways that rendered the suggested edits moot, and we hope the reviewer will appreciate the final result in any case.

::: callout
### Organization

### 2022-09-02 JDSSV_manuscript_submitted.Rnw
tweak figure and add pdf
Updated comments and edits

@@

@@

Throughout the pandemic, both in choropleth maps and time-series graphs, creators have had to grapple with scale. When scales automatically adjust, mental comparisons across time can become misleading, because the colour scale from yesterday does not mean the same thing as it did today, as the range has expanded and the colours have stayed the same. 
It is not that often that day-to-day design changes (or lack thereof) cause fundamental problems with the interpretation of data visualisation, but during the COVID 19 pandemic this situation was both extremely common and had real consequences: people would glance at the current COVID risk map and assume that "red" meant the same thing that it did yesterday, despite the fact that the underlying values had changed\citep{abiadColorDiagramWHOLebanon2020,rubelHeyUpshotNYTCurious2020,matthewsLessonFutzingData2020}. 
One example of this issue over two months of NY Times graphics is shown in \Cref{fig:nyt-colors}; the overall color scheme used changes once, but the variable displayed changes (at least) 4 times over the course of a two month period.

\begin{figure}
\centering
\begin{subfigure}[c]{.25\textwidth}
\includegraphics[width=\textwidth]{nyt-wayback-20200616}
\caption{June 16, 2020}\label{fig:nyt-colors1}
\end{subfigure}\hfill
\begin{subfigure}[c]{.25\textwidth}
\includegraphics[width=\textwidth]{nyt-wayback-20200716}
\caption{July 16, 2020}\label{fig:nyt-colors2}
\end{subfigure}\hfill
\begin{subfigure}[c]{.25\textwidth}
\includegraphics[width=\textwidth]{nyt-wayback-20200718}
\caption{July 18, 2020}\label{fig:nyt-colors3}
\end{subfigure}\hfill
\begin{subfigure}[c]{.25\textwidth}
\includegraphics[width=\textwidth]{nyt-wayback-20200814}
\caption{August 1, 2020}\label{fig:nyt-colors4}
\end{subfigure}\hfill
\caption{A series of screenshots of the NY Times state-by-state display showing evolving color schemes and even metrics; initially, a discrete scale of two-week change in cases is shown; then, a similar color scheme is used with a different metric. Around July 17, the color scheme and variable shown changed to weekly cases per capita (from "fewer" to "more"), and by August 14, the variable shown changed back again ("falling" to "rising") while the color scheme stayed the same.}\label{fig:nyt-colors}
\end{figure}

@@

However, there are some limitations with this presentation: if we are interested in showing the change in counts over time, viewers must assess the slope of the line, rather than its position. We know that slope judgments are much less accurate than position \citep{clevelandGraphicalPerceptionVisual1987}, even in relatively simple situations; when we add in our ability to assess exponential growth, our perceptual accuracy is even more suspect.

Other time-series representations were more abstract, sacrificing exact data representations for a higher-level summary overview that allowed viewers to make comparisons between regions and points in time without the burden of numerical calculations. Several versions from the New York Times are shown in \Cref{fig:sparklines-heatmap-nyt}. The advantage of these displays is that they are very simple and allow for viewers to gain an intuitive understanding of the data, however, they do not present precise numerical information and even the color scale is fairly opaque -- it would be extremely difficult to translate the visualization into any sort of accurate numerical estimate. In addition, by hiding these details from the user, it is very difficult to identify when the representation changes over time, as it did several times during the summer of 2020. The use of a similar color scheme as well as similar geometric representations made it extremely difficult for users to identify that the value represented had changed. Clearly, the purpose of these charts is not to provide numerical precision, but rather a comparative assessment of the status of one region relative to others as well as relative to previous time points.

\begin{figure}
\centering
\includegraphics[width=.3\linewidth]{NYT-heatmap-crop}\hfill
\includegraphics[width=.3\linewidth]{NYT-heatmap-crop3}\hfill
\includegraphics[width=.3\linewidth]{NYT-heatmap-crop2}
\caption{Sparklines style heatmaps embedded in a table of coronavirus data by state. Over the course of Summer 2020, the specific measures shown changed, even though the form of the data remained highly similar. Source: \url{https://www.nytimes.com/interactive/2020/us/coronavirus-us-cases.html} (May 6, 2020; June 15, 2020; and July 24, 2020, respectively) }

\label{fig:sparklines-heatmap-nyt}
\end{figure}

@@

There are many different ways that designers can use time-series data to provide additional contextual information and facilitate comparisons. \Cref{fig:ft-waves} shows covid cases, positivity rate, and hospital admissions from South Africa's Gauteng province, which was one of the first areas to experience a large wave of the Omicron variant. This chart provides context for the Omicron wave, showing that while infections are occurring at a much faster rate than in previous waves, hospital admissions seem to be approximately following previous trends.

\begin{figure}\centering
\includegraphics[width=\linewidth]{ft-waves}\\
\includegraphics[width=\linewidth]{ft-waves-linear}
\caption{Covid Cases, positivity rate, and hospital admissions from South Africa's Gauteng province, on a log scale (top) and linear scale (bottom). Successive waves are overlaid, showing that the increase in cases due to the Omicron variant was sharper than any previously measured increase, while hospitalizations were much more comparable to previous waves. Source: John Burn-Murdoch, \url{https://twitter.com/jburnmurdoch/status/1466480113487392769}}
\label{fig:ft-waves}
\end{figure}

@@

The New York Times intentionally violated typical design guidelines in a controversial chart; the x-axis was wrapped around an Archimedean spiral to provide a sense of periodicity in the year-by-year evolution of case counts, as shown in \Cref{fig:nyt-spiral}. The original form of this chart is prone to the line-width illusion \citep{vanderplasSignsSineIllusion2015}, in addition to the well-known problems with polar charts \citep{hofmannGraphicalTestsPower2012,waldnerComparisonRadialLinear2020}. A re-envisioned version created by Sourya Shrestha, aligns counts on the outside of the spiral, and displays them as radial lines, mitigating some of the issues with the line-width illusion by directly showing each line as its own entity, facilitating direct comparisons. In addition, the re-envisioned chart includes a reference line at 100K cases/day, which allows the reader to compare the severity of different waves directly. Still, it is more work to interpret and compare peaks on this chart than on a similar linear time series chart with the same data, as the reader must assess line width and then do a mental rotation and shift operation in order to compare with any other time period. By design, this type of chart devotes less area to previous years of the pandemic, which may decrease the amount of focus given to that data; whether this is an advantage or disadvantage depends on your perspective and goals when using the chart.

@@

A slightly different approach by the BBC's Visual and Data Journalism team, shown in \Cref{fig:bbc-flower}, displayed the Covid cases and deaths as a growing flower, with the stem proportional to the number of cases and the flower petals showing the deaths over time\citep{thebbcvisualanddatajournalismteamCoronavirusHowCan2020}. This visualization is intended to evoke emotion, with the flower representation as an explicitly recognized symbol of grief; it is clearly not about showing specific case counts and deaths numerically. The visualization comes with sound as well, an attempt to make the pandemic an auditory experience as well as a visual one.

@@

The plague is always the others: This is the short formula for dealing with infectious diseases from a historical perspective \citep{thiessen2021}. Sociologists have coined the term ``othering'' for such attributions of the others \citep{mountz2009}. This refers to the observation that above all new, unknown threats are projected onto ``strangers'' and ``the others''. Closing the borders and restricting access to the country became a popular means of controlling the spread of the disease at several different points during the pandemic. Unsurprisingly, the prevailing visual narrative focused on comparisons often fueled by political rivalry, historical dependencies, recent withdrawal from supranational institutions or regional competitions. An ongoing debate about the true extent of the dangers of covid-19 and how best to combat it, combined with daily availability and public access to data across administrative levels, fostered ongoing competition and the use of leaderboards to show how bad case counts were somewhere else.

@@

The bar charts used for this purpose (see top image in Fig.~\ref{fig:ft-death-ranking}) are straightforward and provide a good way to show the ``leading'' countries or regions. Emphasis on specific countries is either preconfigured by the creator of the chart or can be modified interactively in the online version by check boxes and drop-down menus. These bar charts also allow easy adjustment of the count data to population size or other meaningful standard. Streamgraphs (see lower image in Fig.~\ref{fig:ft-death-ranking}) are a special type of stacked area graphs and became quite popular around 2008 when they were used by the New York Times to visualise box office results of movies \citep{streamgraph}. Their usefulness depends heavily on the clarity of the pattern. Various design choices such as the ordering of the different groups have a strong impact on the visual impression and message conveyed by the graph. Small variations in the proportions of the groups are almost impossible to detect, but streamgraphs give a good indication of which group is predominant at any given time, even though they are subject to the line-width or sine illusion \citep{vanderplasSignsSineIllusion2015}.

@@

\section{To log or not to log}
As COVID cases grow quasi-exponentially while there are susceptible members of the population (subject to the effectiveness of mitigation measures and testing availability), it seems natural to use log scales to allow for more effective comparisons of slight changes in case counts over time. In addition, log scales make it possible to compare regions with different populations or infection rates in the same chart. As noted previously, however, interpreting log scales requires levels of numerical sophistication that may not be appropriate for the general public. Even researchers do not always read and interpret log scales correctly\citep{mengeLogarithmicScalesEcological2018}; expecting the general public to do so is difficult under normal circumstances\citep{hecklerStudentAccuracyReading2013} is difficult. When panic, fear, uncertainty, and doubt about the situation are added to the mix, it is easy to imagine that we become even worse when interpreting graphics.
% Discussion of how this section's conclusions change with the pandemic

One issue with assessing the use of log scales is that their effectiveness changes with the stage of the pandemic and the amount (and varieties) of data shown. Initially, log scales were incredibly useful at showing case counts, because minimal mitigation measures were in place and the growth of case counts (or presumptive positive cases, in absence of available testing) was fairly close to exponential. In addition, the use of log scales allowed for the comparison of nominal cases across entities with large population differences: in the US, we could compare cases in New York and California with cases in Michigan and Washington, even though the population of Michigan and Washington are much lower than the population of either New York or California.

>>>>

In other areas, such as visualization of uncertainty, the pandemic presents new challenges for scientific communication. How do we effectively show the uncertainty in the data amid flawed tests, new strains, changing diagnostic criteria, and incredible variability in global public health infrastructure? Moreover, how do we do this without further decreasing public trust in institutions responsible for shepharding us through the pandemic? While most designers chose to side-step these questions in the moment, as we assess the pandemic in graphical form, we should also consider what was omitted from our visual renderings.

@@

Finally, it is critical that data communication take into account the intended audience. Graphics created to communicate with public health officials should be different from those created for the purpose of informing the public. Designers must carefully consider the visual capabilities, attention constraints, and mathematical sophistication of the audience when creating a chart, but in this era of instantaneous global communication, charts designed for specific audiences may still reach a large population of viewers with different capabilities. In this paper, we focus primarily on graphics which are intended for mass consumption, if only to reduce the scope to something which can be (non-exhaustively) covered in a paper.

@@

Throughout the pandemic, both in choropleth maps and time-series graphs, creators have had to grapple with scale. When scales automatically adjust, mental comparisons across time can become misleading, because the colour scale from yesterday does not mean the same thing as it did today, as the range has expanded and the colours have stayed the same. 
It is not that often that day-to-day design changes (or lack thereof) cause fundamental problems with the interpretation of data visualisation, but during the COVID 19 pandemic this situation was both extremely common and had real consequences: people would glance at the current COVID risk map and assume that "red" meant the same thing that it did yesterday, despite the fact that the underlying values had changed\citep{abiadColorDiagramWHOLebanon2020,rubelHeyUpshotNYTCurious2020,matthewsLessonFutzingData2020}. 
One example of this issue over two months of NY Times graphics is shown in \Cref{fig:sparklines-heatmap-nyt}; the overall color scheme used changes once, but the variable displayed changes (at least) 4 times over the course of a two month period.

@@

However, there are some limitations with this presentation: if we are interested in showing the change in counts over time, viewers must assess the slope of the line, rather than its position. We know that slope judgments are much less accurate than position \citep{clevelandGraphicalPerceptionVisual1987}, even in relatively simple situations; when we add in our ability to assess exponential growth, our perceptual accuracy is even more suspect. This issue will be discussed further in \Cref{sec:logs} when we discuss log scales, which are one solution to our misperception of exponential growth.

Yet another approach to showing the temporal component of the pandemic was to leverage web graphics to show the temporal component through animation; this freed up the $x$ axis which in the default configuration would have been devoted to time and allowed another variable to be displayed. While this approach was common on social media, it was reserved for specific use-cases and more carefully selected variables by professional media outlets. \Cref{fig:bubble-ft} shows an example from the Financial Times which uses the x-axis to show GDP per capita and the y-axis to show vaccination rate, bubble size indicates a country's population. With so many variables and ratios of variables, these charts are complex to read and use, even though they are visually appealing.

\begin{figure}
\centering
\includegraphics{ft_vaccine_bubble}
\caption{Animated bubble chart showing vaccination rate relative to per capita GDP for each country, from the Financial Times (Source: \url{https://ig.ft.com/coronavirus-vaccine-tracker})}\label{fig:bubble-ft}
\end{figure}

Other time-series representations were more abstract, sacrificing exact data representations for a higher-level summary comparing regions and points in time without the burden of numerical calculations. Several versions from the New York Times are shown in \Cref{fig:sparklines-heatmap-nyt}. The advantage of these displays is that they are very simple and allow for viewers to gain an intuitive understanding of the data, however, they do not present precise numerical information and even the color scale is (perhaps intentionally) opaque -- it would be extremely difficult to translate the visualization into any sort of accurate numerical estimate. In addition, by hiding these details from the user, it is very difficult to identify when the underlying mathematical representation changes over time, as it did several times during the summer of 2020. The use of a similar color scheme as well as similar geometric representations made it extremely difficult for users to identify that the value represented had changed. Clearly, the purpose of these charts is not to provide numerical precision, but rather a comparative assessment of the status of one region relative to others as well as relative to previous time points.

% 
% \begin{figure}
% \centering
% \includegraphics[width=.3\linewidth]{NYT-heatmap-crop}\hfill
% \includegraphics[width=.3\linewidth]{NYT-heatmap-crop3}\hfill
% \includegraphics[width=.3\linewidth]{NYT-heatmap-crop2}
% \caption{Sparklines style heatmaps embedded in a table of coronavirus data by state. Over the course of Summer 2020, the specific measures shown changed, even though the form of the data remained highly similar. Source: \url{https://www.nytimes.com/interactive/2020/us/coronavirus-us-cases.html} (May 6, 2020; June 15, 2020; and July 24, 2020, respectively) }
% 
% \label{fig:sparklines-heatmap-nyt}
% \end{figure}

\begin{figure}
\centering
\begin{subfigure}[t]{.25\textwidth}
\includegraphics[width=\textwidth]{nyt-wayback-20200616}
\caption{June 16, 2020}\label{fig:nyt-colors1}
\end{subfigure}\hfill
\begin{subfigure}[t]{.25\textwidth}
\includegraphics[width=\textwidth]{nyt-wayback-20200716}
\caption{July 16, 2020}\label{fig:nyt-colors2}
\end{subfigure}\hfill
\begin{subfigure}[t]{.25\textwidth}
\includegraphics[width=\textwidth]{nyt-wayback-20200718}
\caption{July 18, 2020}\label{fig:nyt-colors3}
\end{subfigure}\hfill
\begin{subfigure}[t]{.25\textwidth}
\includegraphics[width=\textwidth]{nyt-wayback-20200814}
\caption{August 1, 2020}\label{fig:nyt-colors4}
\end{subfigure}\hfill
\caption{A series of screenshots of the NY Times state-by-state display showing evolving color schemes and even metrics; initially, a discrete scale of two-week change in cases is shown; then, a similar color scheme is used with a different metric. Around July 17, the color scheme and variable shown changed to weekly cases per capita (from "fewer" to "more"), and by August 14, the variable shown changed back again ("falling" to "rising") while the color scheme stayed the same.}\label{fig:sparklines-heatmap-nyt}
\end{figure}

@@

There are many different ways that designers can use time-series data to provide additional contextual information and facilitate comparisons. \Cref{fig:ft-waves} shows covid cases, positivity rate, and hospital admissions from South Africa's Gauteng province, which was one of the first areas to experience a large wave of the Omicron variant. This chart provides context for the Omicron wave, showing that while infections are occurring at a much faster rate than in previous waves, hospital admissions seem to be approximately following previous trends. In addition, it is clear that the chart is highlighting the Omicron variant from the colour selection: the red used to show the Omicron numbers is a sharp contrast from the green and blue shades used to show the first three waves of the pandemic. Notably, even though red and green in combination are usually to be avoided for the sake of colourblind viewers, the shades of blue/green variants which are chosen are light enough that the red stands out even to someone with colour vision deficiency.

\begin{figure}\centering
\includegraphics[width=\linewidth]{ft-waves}\\
\includegraphics[width=\linewidth]{ft-waves-linear}
\caption{Covid Cases, positivity rate, and hospital admissions from South Africa's Gauteng province, on a log scale (top) and linear scale (bottom). Successive waves are overlaid, showing that the increase in cases due to the Omicron variant was sharper than any previously measured increase, while hospitalizations were much more comparable to previous waves \citep{burn-murdoch_new_2021-1}.}
\label{fig:ft-waves}
\end{figure}

@@

The New York Times intentionally violated typical design guidelines in a controversial chart; the x-axis was wrapped around an Archimedean spiral to provide a sense of periodicity in the year-by-year evolution of case counts, as shown in \Cref{fig:nyt-spiral}. The original form of this chart is prone to the line-width illusion \citep{vanderplasSignsSineIllusion2015}, in addition to the well-known problems with polar charts \citep{hofmannGraphicalTestsPower2012,waldnerComparisonRadialLinear2020}. A re-envisioned version created by Sourya Shrestha, aligns counts on the outside of the spiral, and displays them as radial lines, mitigating some of the issues with the line-width illusion by directly showing each line as its own entity, facilitating direct comparisons.

In addition, the re-envisioned chart includes a reference line at 100K cases/day, which allows the reader to compare the severity of different waves directly. Still, it is more work to interpret and compare peaks on this chart than on a similar linear time series chart with the same data, as the reader must assess line width and then do a mental rotation and shift operation in order to compare with any other time period. By design, this type of chart devotes less area to previous years of the pandemic, which may decrease the amount of focus given to that data.%; whether this is an advantage or disadvantage depends on your perspective and goals when using the chart.

@@

A slightly different approach by the BBC's Visual and Data Journalism team, shown in \Cref{fig:bbc-flower}, displayed the Covid cases and deaths as a growing flower, with the stem proportional to the number of cases and the flower petals showing the deaths over time\citep{thebbcvisualanddatajournalismteamCoronavirusHowCan2020}. The visualization comes with sound as well, an attempt to make the pandemic an auditory experience as well as a visual one. The form, animation, and auditory experience are all indicative of an emotional appeal \citep{kostelnick_re-emergence_2016,dignazio_data_2020}; with the flower representation as an explicitly recognized symbol of grief; it is clearly not about showing specific case counts and deaths numerically.

@@

The plague is always the others: This is the short formula for dealing with infectious diseases from a historical perspective \citep{thiessen2021}. Closing the borders and restricting access to the country became a popular means of controlling the spread of the disease at several different points during the pandemic. Unsurprisingly, the prevailing visual narrative focused on comparisons often fueled by political rivalry, historical dependencies, recent withdrawal from supranational institutions or regional competitions. An ongoing debate about the true extent of the dangers of covid-19 and how best to combat it, combined with daily availability and public access to data across administrative levels, fostered ongoing competition and the use of leaderboards to show how bad case counts were somewhere else.

@@

The bar charts used for this purpose (see top image in \Cref{fig:ft-death-ranking}) are straightforward and provide a good way to show the ``leading'' countries or regions. Emphasis on specific countries is either preconfigured by the creator of the chart or can be modified interactively in the online version by check boxes and drop-down menus. These bar charts also allow easy adjustment of the count data to population size or other meaningful standard. Streamgraphs (see lower image in \Cref{fig:ft-death-ranking}) are a special type of stacked area graphs and became quite popular around 2008 when they were used by the New York Times to visualise box office results of movies \citep{streamgraph}. Their usefulness depends heavily on the clarity of the pattern, as well as design choices such as the ordering of the different groups. Small variations in the proportions of the groups are almost impossible to detect, but streamgraphs give a good indication of which group is predominant at any given time, even though they are subject to the line-width or sine illusion \citep{vanderplasSignsSineIllusion2015}. When paired with the bar charts showing excess mortality, the streamgraph provides temporal context to the instantaneous information in the bar graph.

@@

\section{To log or not to log}\label{sec:logs}
As COVID cases grow quasi-exponentially while there are susceptible members of the population (subject to the effectiveness of mitigation measures and testing availability), it seems natural to use log scales to allow for more effective comparisons of slight changes in case counts over time. In addition, log scales make it possible to compare regions with different populations or infection rates in the same chart. As noted previously, however, interpreting log scales requires levels of numerical sophistication that may not be appropriate for the general public. Even researchers do not always read and interpret log scales correctly\citep{mengeLogarithmicScalesEcological2018}; expecting the general public to do so is difficult under normal circumstances\citep{hecklerStudentAccuracyReading2013} is difficult. When panic, fear, uncertainty, and doubt about the situation are added to the mix, it is easy to imagine that we become even worse when interpreting graphics.
% Discussion of how this section's conclusions change with the pandemic

One issue with assessing the use of log scales is that their effectiveness changes with the stage of the pandemic and the amount and variety of data shown. Initially, log scales were incredibly useful at showing case counts, because minimal mitigation measures were in place and the growth of case counts (or presumptive positive cases, in absence of available testing) was fairly close to exponential. In addition, the use of log scales allowed for the comparison of nominal cases across entities with large population differences: in the US, we could compare cases in New York and California with cases in Michigan and Washington, even though the population of Michigan and Washington are much lower than the population of either New York or California.
